Describing the scale and composition of calls for police service: a replication and extension using open data

### 2022-02-15 paper.Rmd
response time measure update

Second, we extend the analyses used by @ratcliffe2021policing and @lum2021can. This includes a comparison between measures of 'officer time' with and without the response time. While concerns have been raised about the inclusion of the response time in measures of officer time, largely due to low-priority calls not necessitating a rapid response, the precise impact of their usage on the same data is unclear. We also provide a comprehensive account of the variability in the time consumed by different demand classifications (e.g., crime, community, health). These statistics shed light on the degree to which a disproportionately large amount of officer time is consumed by just a small number of calls - insight which can be used to assess how a reduction in demand, or a shift away from calls for police service towards other social services, might free-up officer time. In addition, we report on the spatial and temporal patterning of each demand classification. Distinctions in these patterns across demand types would be indicative of similarly distinct demand-generating mechanisms (e.g., built environment). Such concentrations in time and space might also inform deployment decisions based on officer expertise and experience, and/or collaboration with other services (e.g., mental health professionals).

@@

Recognizing that frequency counts of calls for service do not necessarily reflect the amount of police resource consumed by public demand, raw counts were augmented with the time spent by officers on each of the six classifications of demand. This information permits the calculation of a proportional breakdown of police time consumed by the different demand classifications throughout the year. The raw numbers were reported in a table. A visual representation of call types nested within each demand classification was summarized in a tree map graphic. The time committed by officers was calculated from when the call was dispatched to an officer to when the officer indicated that the event had been closed. Ratcliffe accounted for the number of officers attending each incident when calculating the amount of time consumed. The aggregate time consumed by different demand classifications was reported using a single count of officer shifts over the year.

@@

To describe the scale and composition of emergency police demand across the US more broadly, @lum2021can used CAD data from nine different police agencies. The agencies themselves were anonymized. Individual incidents were categorized into fourteen different demand classifications, namely, alarms, disorders, vice, domestic, follow-ups and service requests, mental, medical, missing persons, violence, interpersonal-other, property, suspicions, traffic, and admin, agency and non-crime. The authors reported a proportional breakdown of call counts and police time spent across the nine agencies according to the fourteen demand classifications, including the average (mean). For each agency, the authors also reported the average time spent on each demand classification in minutes, and the average across the nine agencies. No descriptive statistics on the range or spread of these figures was reported. The 'time spent' measure used was the _time on scene_ (i.e., excluding response time). The authors considered this to be preferable, since low priority calls may have a non-urgent (slow) response due to the nature of the call, and officers may be reassigned to other duties while on an assignment. Lum and colleagues calculated _time on scene_ as the time from the first officer's arrival to when the call was reported complete. In this way, the calculation did not account for multi-officer responses (e.g., back-up). Instead, the authors calculated _time on scene_ as the time from the first officer's arrival to when the call was reported complete. This was justified for a number of reasons, among them, avoiding overinflation of time due to officers 'flocking' to non-serious incidents, and limitations with the CAD data. A proportional breakdown of the disposition (assist only, report written or arrest citation) was provided for each of the fourteen call types. No spatial or temporal findings were reported.

@@

For each incident, the police response time and time on the scene are reported. The former measures the response time of the first officer on-scene. The latter mimics the measure used by @lum2021can, namely, the amount of time which elapses between when the first officer arrives and when the last unit clears the scene. Here, we refer to the sum of the response time and time on scene as the _total deployed time_. Calls which had a _total deployed time_ or _time on scene_ of zero minutes, negative values or missing were excluded. The data permit a distinction between the time spent with and without the response time. As noted, @lum2021can considered the _time on scene_ measure to be preferable. Here, for completeness, we report times with and without the response time.

Each incident has a _calldescription_ variable which describes the nature of the call. Call descriptions involving administrative duties (e.g., "start of shift information"), completely unknown problems, and non-deployment (e.g., "employee call in / time off") were removed. This left 207 unique call description categories. In the interests of parsimony and ease of interpretation these categories were recoded into 99 broader call descriptions.^[We refer readers to the corresponding GitHub repository for further details on how call descriptions were combined (https://anonymous.4open.science/r/demand_detroit-FCD3/README.md). Readers unfamiliar with R code can review the reference table entitled 'categorization summary'.] This included the removal of incidents involving transporting prisoners and executing warrants, which were deemed to have been erroneously flagged as not being officer-initiated. Each call description was then categorized into the six demand type classifications used in @ratcliffe2021policing, namely, crime, community, medical/public health, proactive, quality of life and traffic. 2.8% of incidents were deemed unclassifiable. For each of the six demand classifications an 'other' incident type was generated for those incidents which consumed less than 0.2% of total deployed time during the year. This left a total of 49 unique categories for the reported breakdowns, capturing ~260,000 individual public calls for service in Detroit during 2019. No information on the final disposition of calls is provided in the open data.

@@

The frequency counts and proportional breakdowns for each demand type classification are reported in Table 1. This demonstrates that in Detroit, a considerable proportion of citizen calls for service, and in turn, time spent by officers, is committed to a diverse array of (often non-criminal) issues. Just 44% of emergency calls for service related to crime incidents. These crime incidents consumed 47% of total deployed time throughout the year, although the figure increases to 51% when considering time on scene in isolation. This likely reflects the fact that crime-related calls have higher priority, and in turn, a swifter response time.^[Crime calls for service had the highest proportion of Priority 1 (highest priority) responses at 52%, followed by traffic (24%), quality of life (17%), community (16%), health (14%), and proactive (3%).] The remainder of total time on scene is spent dealing with traffic duties (16%), quality of life incidents (15%), public and mental health (7%), community issues (5%), and proactive policing (4%).

>>>>

Second, we extend the analyses used by @ratcliffe2021policing and @lum2021can. This includes a comparison between measures of officer time with and without the response time. While concerns have been raised about the inclusion of the response time in measures of officer time, largely due to low-priority calls not necessitating a rapid response, the precise impact of their usage on the same data is unclear. We also provide a comprehensive account of the variability in the time consumed by different demand classifications (e.g., crime, community, health). These statistics shed light on the degree to which a disproportionately large amount of officer time is consumed by just a small number of calls - insight which can be used to assess how a reduction in demand, or a shift away from calls for police service towards other social services, might free-up officer time. In addition, we report on the spatial and temporal patterning of each demand classification. Distinctions in these patterns across demand types would be indicative of similarly distinct demand-generating mechanisms (e.g., built environment). Such concentrations in time and space might also inform deployment decisions based on officer expertise and experience, and/or collaboration with other services (e.g., mental health professionals).

@@

Recognizing that frequency counts of calls for service do not necessarily reflect the amount of police resource consumed by public demand, raw counts were augmented with the time spent by officers on each of the six classifications of demand. This information permits the calculation of a proportional breakdown of police time consumed by the different demand classifications throughout the year. The raw numbers were reported in a table. A visual representation of call types nested within each demand classification was summarized in a tree map graphic. The time committed by officers was calculated from when the call was dispatched to an officer to when the officer indicated that the event had been closed. Ratcliffe accounted for the number of officers attending each incident when calculating the amount of time consumed. The aggregate time consumed by different demand classifications was reported using a single count of officer shifts (7.5 hours) over the year.

@@

To describe the scale and composition of emergency police demand across the US more broadly, @lum2021can used CAD data from nine different police agencies. The agencies themselves were anonymized. Individual incidents were categorized into fourteen different demand classifications, namely, alarms, disorders, vice, domestic, follow-ups and service requests, mental, medical, missing persons, violence, interpersonal-other, property, suspicions, traffic, and admin, agency and non-crime. The authors reported a proportional breakdown of call counts and police time spent across the nine agencies according to the fourteen demand classifications, including the average (mean). For each agency, the authors also reported the average time spent on each demand classification in minutes, and the average across the nine agencies. No descriptive statistics on the range or spread of these figures was reported.

The 'time spent' measure used was the _time on scene_ (i.e., excluding the response time). The authors considered this to be preferable, since low priority calls may have a non-urgent (slow) response due to the nature of the call, and officers may be reassigned to other duties while on an assignment. Lum and colleagues calculated _time on scene_ as the time from the first officer's arrival to when the call was reported as resolved. In this way, the calculation did not account for the additional time consumed by multi-officer responses (e.g., back-up). This was justified for a number of reasons, among them, avoiding overinflation of time due to officers 'flocking' to non-serious incidents, and limitations with the CAD data. A proportional breakdown of the disposition (assist only, report written or arrest citation) was provided for each of the fourteen call types. No spatial or temporal findings were reported.

@@

For each incident, the police response time and time on the scene are reported. The former measures the response time from dispatch to when the first officer arrives on-scene.^[In the raw data, response time is referred to as _traveltime_. This figure does not include the intake or dispatch time which we did not deem to be deployment.] The latter mimics the measure used by @lum2021can, namely, the amount of time which elapses between when the first officer arrives and when the last unit clears the scene. Here, we refer to the sum of the response time and time on scene as the _total deployed time_. Calls which had a response time of negative values or missing were dropped from analyses. Calls which had a _time on scene_ of zero minutes, negative values or missing were also excluded. The data permit a distinction between the time spent with and without the response time. As noted, @lum2021can considered the _time on scene_ measure to be preferable. Here, for completeness, we report times with and without the response time.

Each incident has a _calldescription_ variable which describes the nature of the call. Call descriptions involving administrative duties (e.g., "start of shift information"), completely unknown problems, and non-deployment (e.g., "employee call in / time off") were removed. This left 207 unique call description categories. In the interests of parsimony and ease of interpretation these categories were recoded into 99 broader call descriptions.^[We refer readers to the corresponding GitHub repository for further details on how call descriptions were combined (https://anonymous.4open.science/r/demand_detroit-FCD3/README.md). Readers unfamiliar with R code can review the reference table entitled 'categorization summary'.] This included the removal of incidents involving transporting prisoners and executing warrants, which were deemed to have been erroneously flagged as not being officer-initiated. Each call description was then categorized into the six demand type classifications used in @ratcliffe2021policing, namely, crime, community, medical/public health, proactive, quality of life and traffic. 2.8% of calls were deemed unclassifiable. For each of the six demand classifications an 'other' incident type was generated for those incidents which consumed less than 0.2% of total deployed time during the year. This left a total of 49 unique categories for the reported breakdowns, capturing ~260,000 individual public calls for service in Detroit during 2019. No information on the final disposition of calls is provided in the open data.

@@

The frequency counts and proportional breakdowns for each demand type classification are reported in Table 1. This demonstrates that in Detroit, a considerable proportion of citizen calls for service, and in turn, time spent by officers, is committed to a diverse array of (often non-criminal) issues. Just 44% of emergency calls for service related to crime incidents. These crime incidents consume 51% of time on scene throughout the year. The remainder of total time on scene is spent dealing with traffic duties (16%), quality of life incidents (15%), public and mental health (7%), community issues (5%), and proactive policing (4%). The differences in the proportional breakdown between time on scene and total deployment time are marginal.

<!-- These crime incidents consumed 51% of total deployed time throughout the year, although the figure increases to 51% when considering time on scene in isolation. This likely reflects the fact that crime-related calls have higher priority, and in turn, a swifter response time.^[Crime calls for service had the highest proportion of Priority 1 (highest priority) responses at 52%, followed by traffic (24%), quality of life (17%), community (16%), health (14%), and proactive (3%).] -->
